Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-07-07 ms/ms.tex
update and revise -  YR and IK together

%%% Fig - tau summary %%%
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions. The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers). In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI). 
	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}). Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, rather then Apr~11, 2020.
    }
    \label{fig:tau-summary}
\end{figure}

@@

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant). The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}). 
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
The exceptions are Austria, Belgium, Norway, and United Kingdom. % TODO UPDATE

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions (10 of 12) the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
The exceptions are%, as with the comparison to the simpler models, % TODO UPDATE
Switzerland (see below), and Norway. The latter also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings in more detail.

@@

In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}). % TODO CHECK

@@

%%% Fig - Late %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Italy_τ_posterior.pdf}
    \end{subfigure}
  	~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Mar28/Wuhan_τ_posterior.pdf}
    \end{subfigure}
    \\
    \begin{subfigure}{0.45\textwidth} % TODO UPDATE
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Spain_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/France_τ_posterior.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimate $\hat{\tau}$. Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:late}
\end{figure}

@@

Germany...

Sweden...

Denmark...

%%% Fig - Early %%% 
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth} % TODO UPDATE xlim
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Germany_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Sweden_τ_posterior.pdf}
    \end{subfigure}

	\begin{subfigure}{0.45\textwidth} % TODO UPDATE xlim
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Norway_τ_posterior.pdf}
    \end{subfigure}
    \caption{
    \textbf{Early and exact effect of non-pharmaceutical interventions.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:early}
\end{figure}

@@

The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases. These assessments are made a specific number of days after the intervention began, to accommodate for the expected serial interval~\citep{Banholzer2020} (i.e. time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}. % Table 1

@@

%%% Fig - Prediction %%% TODO UPDATE?
\begin{figure}[h]
    \centering
	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only significant for Germany (\autoref{fig:tau-summary}).
An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may have been influenced by increased risk perception due to domestic or international COVID\=/19-related reports.
Indeed, the evidence supports a change in infection dynamics (i.e. a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}. %TODO check

@@

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e. free or fixed model in \autoref{table:WAIC}). %TODO check
It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.

As countries relieve lockdowns and ease restrictions and the number of cases increases, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.

@@

Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$, respectively.
The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic~\citep{Ferretti2020,Thompson2020}.
The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behaviour.
Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or become ill enough to be quarantined.
In either case, they no longer infect other individuals, and therefore effectively become recorved ($R$).
The model is described by the following set of equations:

@@

%%% Model selection %%%
\paragraph*{Model comparison.}
We perform model selection using two methods.
First, we compute WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
\end{aligned}
\end{equation}
where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
Therefore, we exclude from the computation of WAIC chains with mean log-likelihood that is three standard deviations or more from the overall mean. %TODO

We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}), and plot the simulated dynamics~(\autoref{fig:ppc}).
Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.

%%% Source code %%%
\paragraph*{Source code.} 
We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
Samples from the posterior distributions will be deposited on FigShare.

} % small

%% Table - estimated parameters %%% TODO UPDATE
\begin{landscape}
\begin{table}[]
\centering
\input{../figures/Table-estimated-params.tex}
\caption{
\textbf{Parameter estimates for different regions.}
See \autoref{eq:model} for model parameters.
All estimates are posterior medians.
75\% and 95\% credible intervals given for $\tau$, in days.
$\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
}
\label{table:estimated-params}
\end{table}
\end{landscape}

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Oren Kolodny for discussions and comments.
This work was supported in part by the Israel Science Foundation 552/19 and 1399/17.
}

@@

% Table - WAIC %%% TODO UPDATE
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC_tmp.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy. Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC. $*$ and $**$ mark if the difference was smaller or greater than 2, which is a popular significance level for model comparison~\citep{Kass1995}.
}
\label{table:WAIC}
\end{table}

%%% Fig - tau summary Mar 28 %%%
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Mar28.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions estimated up to Mar~28.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions. Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}). Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:tau-summary-mar28}
\end{figure}

%%% Fig Fr-vs-Es %%% TODO UPDATE
\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig-fr-vs-es.pdf}
    \caption{
    \textbf{\covid confirmed cases in France and Spain.}
    Number of cases proportional to population size (as of 2018). 
    Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
    } 
    \label{fig:fig-fr-vs-es}
\end{figure}

%%%% Fig Switzerland %%%
% \begin{figure}[h]
%     \centering
%         \includegraphics[width=0.75\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
%     \caption{
% 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
%     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
%     }
%     \label{fig:Switzerland}
% \end{figure}

@@

@@

	\subcaption{Wuhan, China}    
        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Wuhan_ppc.pdf}
        \includegraphics[width=\textwidth]{../figures/ppc/free/Wuhan_ppc.pdf}

@@

    Markers represent data ($\vec{X}$). Black line represent a smoothing of the data points using a Savitzky-Golay filter. Colored lines represent posterior predictions from a model with fixed $\tau$ in red, and free $\tau$ in blue. These predictions are made by drawing 1,000 samples from the parameter posterior distribution and then generating a daily case count using the SEIR model in \autoref{eq:model}. Note the differences in the y-axis scale.

>>>>

%%% Fig - tau summary %%%
% generated with /python Fig_tau_summary.py 2020-05-26-Apr11/
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}	
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions. The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers). In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI). The confidence intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}). Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather then Apr~11.
    }
    \label{fig:tau-summary}
\end{figure}

@@

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~23, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
% TODO RMSE reference? (Ilia)
% TODO add table for RMSE (Ilia)
% TODO consider dropping WAIC for RMSE (both)
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark. Notably, the data for Sweden and Denmark does not have a peak during the evaluated dates, possibly leading to wide 95\% confidence intervals on $\hat{tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions (10 of 12) the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
The exceptions are Switzerland (see below), and Norway.
The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings in more detail.

@@

In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~3 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).

@@

%%% Fig - Late %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Italy_τ_posterior.pdf}
    \end{subfigure}
  	~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Mar28/Wuhan_τ_posterior.pdf}
    \end{subfigure}
    \\
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Spain_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/France_τ_posterior.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimate $\hat{\tau}$. Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:late}
\end{figure}

@@

Germany... %TODO (Yoav) \autoref{fig:early}

Sweden and Denmark show early effective start of NPIs. However, for both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates. We suspect this increases the uncertainty in our inference.
Accordingly, WAIC of our model with a free $\tau$ parameter was higher (less fit) than the WAIC of a model with $\tau$ fixed at $\tau^*$ the official date.

%%% Fig - Early %%% 
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Germany_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Sweden_τ_posterior.pdf}
    \end{subfigure}

	\begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Norway_τ_posterior.pdf}
    \end{subfigure}
    \caption{
    \textbf{Early and exact effect of non-pharmaceutical interventions.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:early}
\end{figure}

@@

The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases. These assessments are made a specific number of days after the intervention began, to accommodate for the expected serial interval~\citep{Banholzer2020} (i.e. time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}. % Gatto2020 Table 1

@@

%%% Fig - Prediction
\begin{figure}[h]
    \centering
	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only significant for Germany (\autoref{fig:tau-summary}).
An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may be influenced by increased risk perception due to domestic or international COVID\=/19-related reports~\citep{Arthur2020}.
Indeed, the evidence supports a change in infection dynamics (i.e. a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.

@@

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in most regions (i.e. free or fixed model in \autoref{table:RMSE} and \autoref{table:WAIC}). 
It may be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.

% UPDATE
As countries relieve lockdowns and ease restrictions and the number of cases increases, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.

@@

Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$, respectively.
The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic~\citep{Ferretti2020,Thompson2020}.
The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behaviour.
Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or become ill enough to be quarantined.
In either case, they no longer infect other individuals, and therefore effectively become recovered ($R$).
The model is described by the following set of equations:

@@

%%% Model selection %%%
\paragraph*{Model comparison.}
We perform model selection using two methods.
First, we compute WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
\end{aligned}
\end{equation}
where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
Therefore, we exclude from the computation of WAIC chains with mean log-likelihood that is three standard deviations or more from the overall mean.

We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$ fitted to data up to Apr~11, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}) from Apr~12 to Apr~24, and plot the predicted dynamics~(\autoref{fig:ppc}).
Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
We also compute the expected posterior root mean squared error of these predictions~(\autoref{table:RMSE}).

%%% Source code %%%
\paragraph*{Source code.} 
We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
Samples from the posterior distributions will be deposited on FigShare.
} % small

%% Table - estimated parameters
% generated with /python Table_estimated_params.py 2020-05-26-Apr11/
\begin{landscape}
\begin{table}[]
\centering
\input{../figures/Table-estimated-params.tex}
\caption{
\textbf{Parameter estimates for different regions.}
See \autoref{eq:model} for model parameters.
All estimates are posterior medians.
75\% and 95\% credible intervals given for $\tau$, in days.
$\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
}
\label{table:estimated-params}
\end{table}
\end{landscape}

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Oren Kolodny for discussions and comments.
This work was supported in part by the Israel Science Foundation 3811/19 and 1399/17.
}

@@

% Table - WAIC %%% TODO regenerate table
% generated with /python Table-WAIC.py/
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular significance level for model comparison~\citep{Kass1995}. Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
}
\label{table:WAIC}
\end{table}

% Table - RMSE %%% TODO regenerate table
% generated with /python Table-RMSE.py/
%\begin{table}[h]
%\centering
%\pgfplotstabletypeset[
%    col sep=comma,
%    string type,
%    every head row/.style={before row=\hline,after row=\hline},
%    every last row/.style={after row=\hline},
%    ]{../figures/Table-RMSE.csv}
%\caption{
%\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
%	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
%	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
%	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
%	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model.
%}
%\label{table:RMSE}
%\end{table}

%%% Fig - tau summary Mar 28 %%%
% generated with
% > python Fig_tau_summary.py 2020-06-23-Mar28
% > mv ../figures/Fig-tau-summary.pdf ../figures/Fig-summary-Mar28.pdf
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Mar28.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions estimated up to Mar~28.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions. Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}). Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:tau-summary-mar28}
\end{figure}

%%% Fig Fr-vs-Es
% generated with Fig-fr-vs-es.ipynb
\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig-fr-vs-es.pdf}
    \caption{
    \textbf{\covid confirmed cases in France and Spain.}
    Number of cases proportional to population size (as of 2018). 
    Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
    } 
    \label{fig:fig-fr-vs-es}
\end{figure}

@@

% generated with ppc.sh

@@

%    %%%  
%	\subcaption{Wuhan, China}    
%    \begin{subfigure}{0.45\textwidth}
%        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Wuhan_ppc.pdf}
%    \end{subfigure}
%    ~
%    \begin{subfigure}{0.45\textwidth}
%        \includegraphics[width=\textwidth]{../figures/ppc/free/Wuhan_ppc.pdf}
%    \end{subfigure}

@@

    Markers represent data ($\vec{X}$): circles for data up to Apr~11, 2020, and stars after Apr~12; Apr~11 is represented using a vertical dashed line. Black line represents a smoothing of the data points using a Savitzky-Golay filter. Coloured lines represent posterior predictions from a model with fixed $\tau$ in red, and free $\tau$ in blue. The models were fitted with data up to Apr~11. The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model in \autoref{eq:model}. Note the differences in the y-axis scale.
